---
title: 'Johns Hopkins University Coursera: Regression Models (Project 1: 2015 Februray)'
output: proj1_mpg_regMdl.pdf
---
<br> </br>

## Executive Summary

MPG difference between automatic and manual transmission is statistically analyzed for the 3 predictors of vehicular *weight*, *displacement*, and *horsepower* based on the linear regression models.  For each of the 3 predictors, both summary and regression plot are shown, where data points for automatic and manual transmission are colored 'blue' and 'red' respectively.  The conclusions are:

* MPG difference between automatic and manual transmission is insignifant when the predictor is *weight* because the respective correlations with MPG are nearly the same with -0.98 and -0.97.
* Among the 3 predictors, vehicular *weight* ranks the best, based on its consistently low sigma and high absolute correlation. 
<br> </br>

## Overview

This course project for Regression Models by Johns Hopkins University Coursera is to analyze the [mtcars][mtcars_doc] data (included in the the base version of R) using linear regression models.

### Assignment Context

You work for Motor Trend, a magazine about the automobile industry. Looking at a data set of a collection of cars, they are interested in exploring the relationship between a set of variables and miles per gallon (MPG) (outcome). They are particularly interested in the following two questions:

* “Is an automatic or manual transmission better for MPG”
* "Quantify the MPG difference between automatic and manual transmissions"

### Assignment Question

Take the [mtcars][mtcars_doc] data set and write up an analysis to answer their question using regression models and exploratory data analyses.

Your report must be:

* Written as a PDF printout of a compiled (using knitr) R markdown document.
* Brief. Roughly the equivalent of 2 pages or less for the main text. Supporting figures in an appendix can be included up to 5 total pages including the 2 for the main report. The appendix can only include figures.
* Include a first paragraph executive summary.
<br> </br>

## Summary of Data Preparation and Analysis

Detailed preliminary analyses (not shown for brevity) indicate the use of 3 predictors of *disp*, *hp*, and *wt* for predicting *mpg* by *am* of automatic (0) or manual (1) transmission.  The analyses below start with the following data reduction:

```{r}
# initialize
library( datasets )
str( mtcars )

# reduce mtcars to a subframe consisting of mpg, disp, hp, wt, and am
subfm <- with( mtcars , data.frame( MPG = mpg , DSP = disp , HP = hp , WT = wt , TRNSM = am ) )
```

On the reduced subframe, linear regressions are first applied per predictor based only on TRNSM or *am* = 0 or 1.  Correlation of Coefficients are examined for their closeness.  The closer the values are, the less MPG depends on the vehicular TRNSM.

The plots are then generated for each of the predictor showing points of both TRNSM in 'blue' and 'red' for *am* = 0 and 1, respectively.  The resulting linear regression line is also superimposed.
<br> </br>

## Summary of Linear Regressions

Linear regressions are generated for MPG vs *displacement*, *horsepower*, and *weight* as follows:

**1. MPG vs Displacement Regression**

Three sets of linear regression are generated for *am* = 0, 1, and both:
```{r}
fitDSPa <- lm( MPG ~ DSP , subset( subfm , TRNSM == 0 ) )
fitDSPm <- lm( MPG ~ DSP , subset( subfm , TRNSM == 1 ) )
fitDSP <- lm( MPG ~ DSP , subfm )
smryDSP <- summary( fitDSP , correlation = TRUE )
```

Correlation of Coefficients for *am* = 0 and 1:
```{r}
corDSPa <- summary( fitDSPa , correlation = TRUE )$correlation[ 2 , 1 ]
corDSPm <- summary( fitDSPm , correlation = TRUE )$correlation[ 2 , 1 ]
```

**2. MPG vs Horsepower Regression**

Three sets of linear regression are generated for *am* = 0, 1, and both:
```{r}
fitHPa <- lm( MPG ~ HP , subset( subfm , TRNSM == 0 ) )
fitHPm <- lm( MPG ~ HP , subset( subfm , TRNSM == 1 ) )
fitHP <- lm( MPG ~ HP , subfm )
smryHP <- summary( fitHP , correlation = TRUE )
```

Correlation of Coefficients for *am* = 0 and 1:
```{r}
corHPa <- summary( fitHPa , correlation = TRUE )$correlation[ 2 , 1 ]
corHPm <- summary( fitHPm , correlation = TRUE )$correlation[ 2 , 1 ]
```

**3. MPG vs Weight Regression**

Three sets of linear regression are generated for *am* = 0, 1, and both:
```{r}
fitWTa <- lm( MPG ~ WT , subset( subfm , TRNSM == 0 ) )
fitWTm <- lm( MPG ~ WT , subset( subfm , TRNSM == 1 ) )
fitWT <- lm( MPG ~ WT , subfm )
smryWT <- summary( fitWT , correlation = TRUE )
```

Correlation of Coefficients for *am* = 0 and 1:
```{r}
corWTa <- summary( fitWTa , correlation = TRUE )$correlation[ 2 , 1 ]
corWTm <- summary( fitWTm , correlation = TRUE )$correlation[ 2 , 1 ]
```
<br> </br>

## Summary of Sigmas and Correlations

The following summary table shows the predictor *WT* having the lowest absolute *correlation difference* between *am* = 0 and 2, the lowest *sigma* and the highest absolute *correlation* between *MPG* and predictor:

```{r}
prj1Smry <- matrix( , nrow = 3 , ncol = 3 )
colnames( prj1Smry ) <- c( '|cor(am=0)-cor(am=1)|' , '       sigma' , '|correlation|' )
rownames( prj1Smry ) <- c( 'MPG vs CYL' , 'MPG vs HP' , 'MPG vs WT' )

prj1Smry[ 1 , ] <- c( abs( corDSPa - corDSPm ) , smryDSP$sigma ,  abs( smryDSP$correlation[ 2 , 1 ] ) )
prj1Smry[ 2 , ] <- c( abs( corHPa - corHPm ) , smryHP$sigma ,  abs( smryHP$correlation[ 2 , 1 ] ) )
prj1Smry[ 3 , ] <- c( abs( corWTa - corWTm ) , smryWT$sigma ,  abs( smryWT$correlation[ 2 , 1 ] ) )
round( prj1Smry , 3 )  # round to 3 digits
```

Thus, the vehicular *weight* is a good predictor of its *MPG* irrespective of *transmission* type. 

\pagebreak
## Appendix: Regression Models and Plots

### 1. MPG vs Displacement Regression

Correlation of Coefficients for *am* = 0, 1, and their absolute difference:
```{r}
corDSPa  # correlation of mpg with dsp for am = 0
corDSPm  # correlation of mpg with dsp for am = 1

abs( corDSPa - corDSPm )  # absolute difference

abs( smryDSP$correlation[ 2 , 1 ] )  # absolute correlation for all am
```


```{r , echo=FALSE }
par( mfrow = c( 1 , 1 ) )
with( subfm , plot( DSP , MPG , xlab = 'DSP (cubic inches)' , ylab = 'MPG' , type = 'n' ) )
with( subset( subfm , TRNSM == 0 ) , points( DSP , MPG , col = 'blue' ) )
with( subset( subfm , TRNSM == 1 ) , points( DSP , MPG , col = 'red' ) )
abline( fitDSP , lwd = 2 )
legend( 'topright' , pch = 1 , col = c( 'blue' , 'red' ) ,
  legend = c( 'Auto' , 'Manual' ) )
title( 'MPG by DSP' )
text( x = 470 , y = seq( from = 28 , to = 25 , by = -2 ) , adj = c( 1 , 1 ) ,
  labels = c( paste( 'sigma:' , round( smryDSP$sigma , digits = 3 ) ) ,
  paste( 'correlation:' , round( smryDSP$correlation[ 2 , 1 ] , digits = 3 ) ) ) )
```
Figure 1.  Regression plot of *mpg* with *dsp* for both *am*.


### 2. MPG vs Horsepower Regression

Correlation of Coefficients for *am* = 0, 1, and their absolute difference:
```{r}
corHPa  # correlation of mpg with hp for am = 0
corHPm  # correlation of mpg with hp for am = 1

abs( corHPa - corHPm )  # absolute difference

abs( smryHP$correlation[ 2 , 1 ] )  # absolute correlation for all am
```


```{r , echo=FALSE }
par( mfrow = c( 1 , 1 ) )
with( subfm , plot( HP , MPG , xlab = 'HP (hp)' , ylab = 'MPG' , type = 'n' ) )
with( subset( subfm , TRNSM == 0 ) , points( HP , MPG , col = 'blue' ) )
with( subset( subfm , TRNSM == 1 ) , points( HP , MPG , col = 'red' ) )
abline( fitHP , lwd = 2 )
legend( 'topright' , pch = 1 , col = c( 'blue' , 'red' ) ,
  legend = c( 'Auto' , 'Manual' ) )
title( 'MPG by HP' )
text( x = 335 , y = seq( from = 28 , to = 25 , by = -2 ) , adj = c( 1 , 1 ) ,
  labels = c( paste( 'sigma:' , round( smryHP$sigma , digits = 3 ) ) ,
  paste( 'correlation:' , round( smryHP$correlation[ 2 , 1 ] , digits = 3 ) ) ) )
```
Figure 2.  Regression plot of *mpg* with *hp* for both *am*.


### 3. MPG vs Weight Regression

Correlation of Coefficients for *am* = 0, 1, and their absolute difference:
```{r}
corWTa  # correlation of mpg with wt for am = 0
corWTm  # correlation of mpg with wt for am = 1

abs( corWTa - corWTm )  # absolute difference

abs( smryWT$correlation[ 2 , 1 ] )  # absolute correlation for all am
```


```{r , echo=FALSE }
par( mfrow = c( 1 , 1 ) )
with( subfm , plot( WT , MPG , xlab = 'WT (klb)' , ylab = 'MPG' , type = 'n' ) )
with( subset( subfm , TRNSM == 0 ) , points( WT , MPG , col = 'blue' ) )
with( subset( subfm , TRNSM == 1 ) , points( WT , MPG , col = 'red' ) )
abline( fitWT , lwd = 2 )
legend( 'topright' , pch = 1 , col = c( 'blue' , 'red' ) ,
  legend = c( 'Auto' , 'Manual' ) )
title( 'MPG by WT' )
text( x = 5.4 , y = seq( from = 28 , to = 25 , by = -2 ) , adj = c( 1 , 1 ) ,
  labels = c( paste( 'sigma:' , round( smryWT$sigma , digits = 3 ) ) ,
  paste( 'correlation:' , round( smryWT$correlation[ 2 , 1 ] , digits = 3 ) ) ) )
```
Figure 3.  Regression plot of *mpg* with *wt* for both *am*.
